\documentclass[a4paper]{tufte-book}

\title{The Parachute Book\newline Vol I: Bare Metal}
\author{ Matt Gumbley, DevZendo.org }

%\date % without \date command, current date is supplied

%\geometry{showframe} % display margins for debugging page layout


% The tufte-latex package is great for pdflatex, but TeXPad uses xelatex, which fouls up the header, and doesn't build unless the following section is added. The header doesn't use a smallcaps font, and I had to change the main text and math font to make it look a bit better.

% Thanks to https://tex.stackexchange.com/questions/200722/xetex-seems-to-break-headers-in-tufte-handout
\usepackage{fontspec}
\usepackage{ifxetex}


\ifxetex
  \newcommand{\textls}[2][5]{%
    \begingroup\addfontfeatures{LetterSpace=#1}#2\endgroup
  }
  \renewcommand{\allcapsspacing}[1]{\textls[15]{#1}}
  \renewcommand{\smallcapsspacing}[1]{\textls[10]{#1}}
  \renewcommand{\allcaps}[1]{\textls[15]{\MakeTextUppercase{#1}}}
  \renewcommand{\smallcaps}[1]{\smallcapsspacing{\scshape\MakeTextLowercase{#1}}}
  \renewcommand{\textsc}[1]{\smallcapsspacing{\textsmallcaps{#1}}}
\fi

\setmainfont[
  Renderer=Basic,
  Numbers=OldStyle,
  Scale=1.0,
  Extension=.otf,
  UprightFont=*-regular,
  ItalicFont=*-italic,
  BoldFont=*-bold,
  BoldItalicFont=*-bolditalic]{texgyrepagella} %% The Palatino from the TeX Gyre Project

% Get the small caps right. Too much fiddling going on when using xelatex and tufte-latex.
% The title and author aren't as nice as with pdflatex, but at least it builds.
% Path = /usr/share/texmf/fonts/opentype/public/tex-gyre/ ,
\setmainfont{texgyrepagella-regular.otf}[
  BoldFont       = texgyrepagella-bold.otf,
  ItalicFont     = texgyrepagella-italic.otf,
  BoldItalicFont = texgyrepagella-bolditalic.otf,
  SmallCapsFeatures={Letters=SmallCaps},
]

% \setsansfont[Renderer=Basic, Scale=0.90]{Helvetica}
% Set up the spacing using fontspec features
\renewcommand\allcapsspacing[1]{{\addfontfeature{LetterSpace=15}#1}}
\renewcommand\smallcapsspacing[1]{{\addfontfeature{LetterSpace=10}#1}}


\usepackage{graphicx} % allow embedded images
\setkeys{Gin}{width=\linewidth,totalheight=\textheight,keepaspectratio}
\graphicspath{{graphics/}} % set of paths to search for images
\usepackage{amsmath}  % extended mathematics
\usepackage{booktabs} % book-quality tables
\usepackage{units}    % non-stacked fractions and better unit spacing
\usepackage{multicol} % multiple column layout facilities
\usepackage{lipsum}   % filler text
\usepackage{fancyvrb} % extended verbatim environments
\usepackage{hyperref} % URLs
\usepackage{xcolor}   % For coloured text
\usepackage{soul}     % For highlighted text
\fvset{fontsize=\normalsize}% default font size for fancy-verbatim environments

% Standardize command font styles and environments
\newcommand{\doccmd}[1]{\texttt{\textbackslash#1}}% command name -- adds backslash automatically
\newcommand{\docopt}[1]{\ensuremath{\langle}\textrm{\textit{#1}}\ensuremath{\rangle}}% optional command argument
\newcommand{\docarg}[1]{\textrm{\textit{#1}}}% (required) command argument
\newcommand{\docenv}[1]{\textsf{#1}}% environment name
\newcommand{\docpkg}[1]{\texttt{#1}}% package name
\newcommand{\doccls}[1]{\texttt{#1}}% document class name
\newcommand{\docclsopt}[1]{\texttt{#1}}% document class option name
\newenvironment{docspec}{\begin{quote}\noindent}{\end{quote}}% command specification environment


\newcommand{\floor}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\ceil}[1]{\left\lceil #1 \right\rceil}

% Generates the index
\usepackage{makeidx}
\makeindex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\frontmatter

\maketitle % this prints the handout title, author, and date

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\begin{fullwidth}
~\vfill
\thispagestyle{empty}
\setlength{\parindent}{0pt}
\setlength{\parskip}{\baselineskip}
Copyright \copyright\ \the\year\ \thanklessauthor

% about DevZendo.org ?


\par\smallcaps{https://devzendo.github.io/parachute}

\par Licensed under the Apache License, Version 2.0 (the ``License''); you may not
use this file except in compliance with the License. You may obtain a copy
of the License at \url{http://www.apache.org/licenses/LICENSE-2.0}. Unless
required by applicable law or agreed to in writing, software distributed
under the License is distributed on an \smallcaps{``AS IS'' BASIS, WITHOUT
WARRANTIES OR CONDITIONS OF ANY KIND}, either express or implied. See the
License for the specific language governing permissions and limitations
under the License.\index{license}

\vspace{5mm}

'Transputer' and 'occam' were trademarks of Inmos Limited, later, SGS-Thompson Microelectronics, later, ST Microelectronics.

\vspace{5mm}

\par\textit{Parachute 0.0.2, June 2026}
\end{fullwidth}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \pagebreak
    \tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pagebreak
~\vfill
\thispagestyle{empty}
\begin{doublespace}
\noindent\fontsize{18}{22}\selectfont\itshape
\nohyphenation

\noindent
In memoriam\\
\vspace{5mm}
Sir Charles Antony Richard Hoare,\\
FRS, FREng.\\
\vspace{5mm}
11 January 1934 - 5 March 2026

\end{doublespace}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


% each chapter opens with a brief paragraph stating its scope and purpose, and closes with a summary and forward pointer.

    
\pagebreak
\part{Preface}
\section{About this book}
    \newthought{Parachute is...} an open, simple and useful personal computing stack — built around the Inmos Transputer (in emulation).\footnote{Hold on. I want to set some expectations here - although I've been working on this for many years, off and on, this system is currently still \textit{a beginning}.} \\
\vspace{5mm}
\vspace{5mm}
        
	It is a long-term project to build a small systems architecture
        from the ground up: emulator, toolchain, OS kernel, and applications — designed
        to run on current operating systems, and cheap, widely available hardware.

\vspace{5mm}
    Small enough that a person can understand the whole system.
\vspace{5mm}

	This book is a tutorial introduction and complete reference to \textit{part} of the Parachute system: its emulator, hardware\footnote{Initially, the Raspberry Pi Pico, currently around £3.80 in the UK, July 2026.} and languages. The focus of this volume is on running single, small - but parallel - applications on it directly, with \textit{no operating system}.\footnote{Hence the 'Bare Metal' of the title of this first volume.}
	
	In its heyday, the Transputer was used either as a network of many microcontrollers in an embedded system, or as a co-processor network for an existing mainstream computer.\footnote{e.g. the B008 PC ISA board, and the Atari Transputer Workstation.} Later developments created operating systems for pure Transputer hardware.\footnote{e.g. the Helios operating system.} 
	
	This volume focuses on the first two applications, using the Parachute emulator instead of actual Transputer hardware.
		
	Volume two - not yet started - will cover the forthcoming microkernel OS of the Parachute system.

\section{Who should read this book}

Those who yearn for a return to simple, distraction-free devices, that won't become obsolete, due to them being buildable from open technology.

Developers and tinkerers who remember when personal computing felt like a frontier — or those who wish they'd been there.

Those wanting to explore true parallel processing, with a language designed to work with it.

People who want to understand their tools - all the way down.

Anyone who thinks the direction of travel in modern systems (more abstraction, more complexity, more magic, less freedom, no right-to-repair) isn't the only possible direction. Perhaps innovative ideas from the past were \textit{before their time}, had much to offer, and are worth re-visiting in a modern setting?

Those who would like to help form a community building this: it's intended to be a community project as much as a technical one. \newthought{There's a lot to build. \textit{Come help}.}


\section{How to read this book}
The book can be read sequentially - each section builds on the previous one.

The next part, \textit{Foundations} on page \pageref{foundations}, describes the vision and motivation behind the project. 

It then gives a high-level description of the Transputer processor, and how our emulator relates to this. There are many differences.\footnote{Appendix \ref{appendix:compatibility} on page \pageref{appendix:compatibility} details how accurately the emulator mirrors the behaviour of real Transputers.}

When used as a co-processor to a host computer, the Transputer made use of host computer I/O facilities to access the filesystem, the real-time clock, the textual console and keyboard. This client/server aspect of Transputers is explained in depth, which leads into a glimpse forward to the forthcoming OS's device driver system.

\vspace{5mm}

After gaining a solid understanding of the architecture provided by Parachute, the next part of the book, \textit{Getting Started} on page \pageref{getting-started} covers downloading, verifying, and installing the software for the various supported platforms, and also flashing the emulator image onto Raspberry Pi Pico boards.

\vspace{5mm}

Once installed, the \textit{Programming} section of the book on page \pageref{tutorial-introduction} gives a variety of small programs, in several languages, that will show you more of the architecture, and how to work effectively with the 'bare metal' emulator.

These examples grow in complexity, and move from single processes, to multiple co-operating processes, to multiple processors in networks.

\vspace{5mm}

Finally the book ends with a complete reference on page \pageref{reference} to the various programs and languages available within the system.

\vspace{5mm}

By the end of the book, you should be able to write small programs for this interesting modern 'take' on an architecture that was once poised to change the world.

\newthought{The future was parallel. \textit{Explore it}.}

\section{What you need to work with this project}
For the desktop version of Parachute, you need a computer with a supported operating system. It does not have to be state-of-the-art.

For working with the embedded emulator, start with a Raspberry Pi Pico. Version one is OK.

\vspace{5mm}

You do not need real Transputer hardware. These appear occasionally on auction sites, but they can be harder to work with in today's systems, requiring subtle interfacing. They are generally considered by the author to be made of \textit{Unobtanium}.

\vspace{5mm}

It's not viable to build a new, popular system from hardware that's difficult to source, and that isn't currently manufactured.\footnote{ The last chips that came from the Transputer architecture, the ST Microelectronics STi5100, ceased in 2016 as ST exited the set-top-box market. https://www.cpushack.com/2016/02/03/the-end-of-the-omega/ } Supply chain breakdowns during the COVID pandemic convinced me of the need to get Parachute running on highly available hardware. By making the emulator portable and cross-platform, I hope it can run on whatever system you have at hand. I chose the Raspberry Pi Pico specifically because it is widely available in a variety of implementations from many vendors, and is inexpensive. Clusters of original Transputers might have been prohibitively expensive for hobbyists back in the day - but clusters of Pi Picos are not.

% TODO work on this - treads on the install requirements section. First person?

% who this is for, what you need, conventions, how chapters relate, how to get the source code
% Who is reading a document?
% What do they want to accomplish?
% Why do they want to accomplish it?
    

\subsection{Conventions used in this book}

The following typographical conventions are used in this book:

\textit{Italic} represents new terms.

\verb|Constant width| is used for code fragments, filenames, text emitted by programs.

\texttt{\textbf{Constant width bold}} is used for input items the user should type.
 

\subsection{Using code examples}

I appreciate, but do not require attribution. This usually takes the form "The Parachute Book, Vol I", Copyright 2026 by Matt Gumbley, DevZendo.org.


\subsection{How to contact us}

Email: \verb|matt.gumbley@devzendo.org| \footnote{I'll set up a \texttt{groups.io} mailing group in the near future.} \\
\vspace{5mm}
Mastodon: \verb|@M0CUV@mastodon.radio| \footnote{For more information on the Mastodon decentralised, non-algorithmic social network, please see \texttt{https://joinmastodon.org}.} \\
\vspace{5mm}
Web: \verb|https://devzendo.github.io/parachute| \\
\vspace{5mm}
Source code: \verb|https://github.com/devzendo/parachute| \footnote{This is the top-level repo where issues should be created - there are several other sub-project repositories.}


\subsection{Policy on contributions and use of Large Language Models ('AI')}
I welcome contributions from humans via pull requests on all aspects of the system. \\
\vspace{5mm}
I use Claude.ai occasionally for \textit{small-scale} assistance with code and prose; for advice, and as a more focused search/analysis tool. \textit{All} its output is vetted thoroughly by me: I am \textit{highly alert} to its potential for hallucination and its propensity to confabulate and mislead. \\
\vspace{5mm}
I gratefully acknowledge the work of others that was stolen without consent in the creation of Large Language Models. I understand the environmental impact of this technology. \\
\vspace{5mm}

The largest use of Claude.ai in this project has been the translation of the text output, commentary and variables of the Small-C compiler and assembler\footnote{Written by Óscar Toledo Gutiérrez.} from Spanish to English. The translation makes sense to me, as someone familiar with compiler internals, and the binary output has been tested with both language versions, and is identical. \\
\vspace{5mm}

Should you wish to submit changes or pull requests to any of the Parachute projects, I \textit{strongly prefer} Test-Driven Development, as distinct from writing tests after the code. Some parts (e.g. the IServer) were written after I adopted TDD, and are mostly test-driven; some parts (e.g. the Emulator core) predate this. When changing the older code, I prefer to get it under test first, then continue using TDD. (e.g. the asynchronous link subsystem.)\\
\vspace{5mm}
\textbf{I will not accept AI-generated slop in contributions}.

\vfill


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\pagebreak
\part{Foundations}
\label{foundations}

\section{Inspiration}
% vision/motivation, what a Transputer is, the architecture, the ecosystem (IServer etc.)

Parachute is motivated by the ideas of people who believed software could be lean, comprehensible, and purposeful:
\begin{itemize}
	\item The Inmos Transputer, occam, and C.A.R. Hoare's CSP — the technical foundation (with other languages)
	\item Niklaus Wirth's Project Oberon and his essay 'A Plea for Lean Software' — a philosophy of restraint
	\item The Amiga OS — elegance and capability in a small footprint
	\item Psion PDAs and the SIBO/EPOC operating systems — productive, focused, astonishingly efficient
	\item The Palm Pilot — proof that constraints produce clarity
\end{itemize}

Think: the early Mac, the Amiga, early Windows — but parallel, distributed, and built on a CPU that was genuinely ahead of its time.
	
And remember the PDA? Pocket-sized. Days of battery life on AAs. Distraction-free and productive. The smartphone was not its natural successor. Touch screens are necessary, but not sufficient for productive work - proper keyboards are essential. 

% each chapter opens with a brief paragraph stating its scope and purpose, and closes with a summary and forward pointer.

\section{Transputer Processor Overview}
This section will introduce the Transputer and its features as presented via the Parachute emulator.
 
\marginnote{This is only the very briefest orientation to what makes a Transputer \textit{different} from other microprocessors. For a much more in-depth overview, I highly recommend the following book which is available from \texttt{https://transputer.net/iset/
isbn-013929134-2/tthb.pdf}: \cite{tth}}

There are several members in the Transputer family of microprocessors, but for simplicity this guide will describe only two of them, as they are implemented by the emulator:
\begin{itemize}
	\item The T425 - a 32-bit, integer-only processor with 4KB internal memory, support for debugging, and 4 links.
	\item The T805 - like the T425, but adds floating-point instructions. 
\end{itemize}

The Parachute emulator is somewhere in-between these two, closer to the T425. Floating support is not yet added, apart from some FP  instructions present in the T425. There was a 16-bit family of Transputers, with fewer links and disk controller hardware, and there was to be a successor to the T8xx, the T9000 - this never went into production.

% memory

A Transputer is different from typical microprocessors in that it had a small amount of in-built memory.\footnote{The T800 had 4KB; Parachute on desktop has by default 4MB but this can be changed via command line flags; Parachute on the Pi Pico has around 200KB - and is currently only making use of a single core of the RP2040: this will change in a later release.} It also provided an external memory interface for connecting more RAM. Each Transputer in a system had access to its own memory - different to today's systems where memory is shared between multiple cores.

% scheduler

The Transputer also contained a microcoded scheduler with two levels of priority. This permitted multitasking without the need for a full operating system.

% links

Multiple Transputers could be connected together by use of several \textit{links} \index{links}, numbered from zero to three - these were two-wire bidirectional serial connections that could be attached to other Transputers to form a network for fast inter-processor communications. Programs and data can be sent over this network to distribute the processing. The T800 could achieve 1.47 Mbytes/second at 20Mbps.\footnote{In the Parachute emulator, hosted on Raspberry Pi Pico, link speed is currently very slow - it's more of a proof of concept, as it does not currently use the Pico's PIO subsystem. A later release will fix this.} 

Typically, several Transputers would be used as parallel controllers in an embedded system, connected together by these links. One link would be used to connect the network of Transputers to a host computer, using a \index{link adapter} \textit{link adapter} chip.  There is no intention to be electrically compatible with real Transputers, either via links, or any other interfacing. The Parachute emulator is a \textit{new Transputer-like module} - it will eventually have fast links that interoperate with other Picos running the emulator - but there's no plan to develop timing-compatible links for attachment to a real T800, say.\footnote{Never say never - it could be done, with appropriate timing and voltage level shifting - it's just not a goal of mine yet.} The signalling protocol used by the Pi Pico GPIO links is the same as that used on real Transputers.

The Parachute emulator provides several different link implementations - not just bidirectional serial over GPIO pins. To interface the Pico-based emulator to a host, link zero is exposed as a USB CDC interface, so you can connect the Pico via USB cable, and run host software that communicates via a serial port. In addition to the emulator for Pico, there is also a separate \textit{link adapter} which has one link over USB CDC serial, and one link over the bidirectional serial GPIO pins.\footnote{Traditional Transputers could be interfaced using a C012 link adapter chip. The USB adapter is the modern equivalent.} Parachute modernises the link subsystem, to make it trivial to connect to a host. In the desktop installation, the link between the emulator and IServer (see next section) is made using a FIFO or Named Pipe. This could be enhanced to be a TCP socket, extending the reach of the system to the Internet.

% accuracy

The emulator isn't \textit{timing accurate} - there's an \textit{emulated clock} in the emulator, and an instruction takes as many emulated clock cycles as the instruction reference says it should - but this does not relate to real time, as an actual Transputer would.

Emulation is therefore, \textit{instruction-level accurate}. 

% csp

The programming model is that of multiple processes, communicating with each other over \textit{channels} \index{channel}. A channel is simply a word of memory, where communication between processes is synchronised. The links are accessed via specific locations in the memory map - to a running program there is no distinction between a channel in memory and one that sends/receives data over a link: the Transputer copies messages between memory channels, or directs the DMA engines in the link subsystems to handle communications externally based on the channel address. The instructions for hardware or memory channels are the same. There is obviously a performance penalty if communication is external - but link communications take place independently in parallel to the CPU executing. Processes involved in communication are descheduled and rescheduled during communication.

This model is based on the \index{CSP} \textit{CSP} (Communicating Sequential Processes) design formalism\footnote{Developed by C.A.R. Hoare.}. When designing computer systems, we often try to force a problem that is inherently concurrent and asynchronous into a sequential substrate. CSP - and the Transputer is a hardware implementation of it - gives you a framework for expressing your problem in its 'natural' state, without decomposing it into a sequential form.

% languages

The Transputer was originally programmed in \texttt{occam} \index{occam} - a language designed to directly support concurrent and multi-processor programming.\footnote{Parachute does not yet include \texttt{occam}, but I intend to add it. For now, we have assembler, Small-C and eForth.}

% risc-y

Machine code on the Transputer is RISC-like, with common instructions taking a single nybble. Less common instructions take more storage to encode via an escaping mechanism. Each instruction is therefore variable length. Our assemblers take care of this, and the generation of the most compact code.

% booting

A real Transputer can be switched to boot from either a ROM or by receiving its boot program down one of its links. The Parachute emulator supports this, and also allows a program image to be loaded directly into its memory and executed, as though it had booted via a link. This is much faster than booting from a link, and is used for several development tools.
 
\section{IServer client/server system}


\section{Project Website}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\pagebreak
\part{Getting Started}
\label{getting-started}
% In the tutorial part, resist the temptation to front-load explanation. The reader should do something that works in the first chapter before you explain why it works. Explanation belongs in Part I or in clearly marked 'under the hood' sections, not interspersed in tutorial steps. This is where many self-authored technical books drift.

\section{Project Website}

The website for the parachute software is located at
\url{https://devzendo.github.io/parachute}. On the website, you'll find
links to our \smallcaps{github} repository, mailing lists, bug tracker, and documentation. 

Releases of parachute may be found in the \smallcaps{releases} area of the \smallcaps{github} repository: \url{https://github.com/devzendo/parachute/releases} .



\section{Downloading}




\section{Verifying downloads}




\section{Installing}
\subsection{Installing on Windows}
\subsection{Installing on macOS}
\subsection{Installing on Linux}
\subsection{Installing on Raspberry Pi Pico}


\section{Uninstalling}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\pagebreak
\part{Programming}

% tutorials in escalating complexity, one concept per chapter, each self-contained but building on prior ones; multiple languages means likely a chapter per language introducing it before the complexity builds up

\section{Tutorial introduction}
\label{tutorial-introduction}
\subsection{Hello World, prebuilt}
\subsection{Assembling Hello World}
\subsection{Building Small-C Hello World}
\subsection{eForth}
\subsection{The Mandlebrot Set}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\pagebreak
\part{Reference}
\label{reference}
% one chapter per tool/language, dense, structured consistently (syntax, semantics, options, error conditions)

\section{Software}
\subsection{Emulator}
\subsection{IServer}
\subsection{eForth}

\section{Languages}
\subsection{Macro Assembler}
\subsection{Small-C Compiler}
\subsection{Small-C Assembler}
\subsection{eForth}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\pagebreak
\appendix
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}
\renewcommand{\thesubsection}{\Alph{subsection}}

% quick reference tables, platform-specific notes, bibliography


\subsection{Emulator Compatibility}
\label{appendix:compatibility}

\subsection{IServer Compatibility}

\subsection{Components of Parachute}

There are several sub-components to Parachute, written over a long time period - in the languages I was studying and working with commercially at the time.

\vspace{5mm}

The emulator and IServer (transputer-emulator) is the oldest component, and is written in C++14, using CMake as its build system. It builds on the three 'desktop' OSes (macOS (Intel, Apple Silicon), Linux (Intel, ARM) and Windows 10 (Intel)). The earlier code here (the emulator) was not written with TDD; the later code (the IServer) was. The emulator and a link adapter also builds for Raspberry Pi Pico.

eForth (transputer-eforth) is written in assembly language, specifically an open source implementation of a subset of Microsoft Macro Assembler 6 that understands the Transputer instruction set.

This macro assembler (transputer-macro-assembler) is written in Scala 2, completely with TDD. It makes use of several other of my repositories (group-parent, group-parent-scala, cross-platform-launcher-plugin, common-code).

The forthcoming C compiler (retro-c-compiler) is written in Rust, completely with TDD. This is in early development and is not yet part of the Parachute distribution.

The Small-C compiler and assembler (transputer-k-r-c-compiler) is written in its own language, Small-C. It currently only builds on a 32-bit OS; I use Debian Bookworm to build it for Linux, then use this to build it again for Transputer.

This book (parachute-book) is written using \LaTeX. It builds on macOS using MacPorts TeXLive, and in a Docker container that provides all the \LaTeX \space tools and packages (documentation-docker).

\vspace{5mm}

I use Apache Maven as an overall orchestrator of the build of all these projects - even though it is typically used for building JVM programs. No other build system understands the concept of a \textit{project lifecycle} in the way Maven does. I haven't found a replacement for its \textit{release} versioning/mechanism. 

Each project contains Maven's \texttt{pom.xml} and other control files to produce the final project artefact.

\subsection{Building Parachute}
The full build of the system as a whole is done on a private Jenkins cluster of virtual machines, to support all the different operating system builds that Parachute is targeting. Of course, this makes it hard for other contributors to build the whole system.

To alleviate this, I have built a Docker image that contains all the tools needed to build everything, together with some scripts that can clone all necessary repositories, and start the build of everything.

This can be found at \verb|https://github.com/devzendo/parachute-docker|.

\hl{INSTRUCTIONS ON HOW TO USE IT}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% The back matter contains appendices, bibliographies, indices, glossaries, etc.

\backmatter


\pagebreak
\bibliography{parachute-book}
\bibliographystyle{plainnat}

\printindex

\end{document}
